---
title: "HW 1"
output:
  html_document: default
  pdf_document: default
tags: concept, strategy
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE,  
                      warning = FALSE, message = FALSE, 
                      fig.align = "center",
                      R.options = list(max.print=100))

# Edit the file starting below
```


### Andrew Manrriquez (ajm7246)

**You will submit this homework assignment as a pdf file on Gradescope.**

*For all questions, include the R commands/functions that you used to find your answer (show R chunk). Answers without supporting code will not receive credit. Write full sentences to describe your findings.*

------------------------------------------------------------------------

## Part 1: (11 pts)

The dataset `mtcars` was extracted from the 1974 Motor Trend US magazine, and comprises fuel consumption and other aspects of automobile design and performance for different cars (1973-74 models). Look up the documentation for this data frame with a description of the variables by typing `?mtcars` **in the console** pane.

### Question 1: (2 pt)

Take a look at the first 6 rows of the dataset by using an `R` function in the code chunk below. Do you know about any (or all) of these cars?

```{r}
# Loading the data set mtcars
data("mtcars")
# Display the first 6 rows of the data set mtcars
head(mtcars)
```

**I know nothing about these cars but I have heard of the brand Mazda.**

------------------------------------------------------------------------

### Question 2: (2 pts)

How many rows and columns are there in this data frame in total?

```{r}
# The number of rows in the data set 
nrow(mtcars)
# The number of columns in the data set
ncol(mtcars)
```

**The `mtcars` data set has 32 rows and 11 columns.**

------------------------------------------------------------------------

### Question 3: (1 pt)

Save `mtcars` in your environment and name it as your `eid`. From now on, use this new object instead of the built-in dataset.

```{r}
# Saving the data set mtcars into the environment under the name ajm7246
ajm7246 = mtcars
```

**`mtcars` has now been saved under the variable name `ajm7246`.**

------------------------------------------------------------------------

### Question 4: (2 pts)

When is your birthday? Using indexing, grab the value of `mpg` that corresponds to the day of your birthday (should be a number between 1 and 31).

```{r}
# Using indexing to grab the 17th value of mpg.
ajm7246$mpg[17]
```

**My birthday is on the 17th and the 17th value of `mpg` corresponds to 14.7.**

------------------------------------------------------------------------

### Question 5: (2 pts)

Using logical indexing, count the number of rows in the dataset where the variable `mpg` takes on values greater than 30.

```{r}
# Creating a new variable that only shows values greater than 30 mpg
greater_30_mpg = ajm7246[ajm7246$mpg >30,]
# Counting the number of rows greater than 30
nrow(greater_30_mpg)
```

**There are four rows in the data set that have a value greater than 30 in the `mpg` variable.**

------------------------------------------------------------------------

### Question 6: (2 pts)

Let's create a new variable called `kpl` which converts the fuel efficiency `mpg` in kilometers per liter. Knowing that 1 mpg corresponds to 0.425 kpl, complete the following code and calculate the max kpl:

```{r}
# Loading the package tidyverse
library(tidyverse)
# Creating a new variable called kpl (converts mpg to kpl)
ajm7246 = ajm7246 %>% mutate(kpl=(ajm7246$mpg*0.425))
# Calculate the max kpl
max(ajm7246$kpl)
```

**After converting `mpg` to `kpl` and adding it as a new variable to the data set, the max `kpl` is calculated to be 14.4075.**

------------------------------------------------------------------------

## Part 2: (6 pts)

Let's quickly explore another built-in dataset: `airquality` which contains information about daily air quality measruements in New York, May to September 1973. 

### Question 7: (2 pts)

Calculate the mean `Ozone` (in ppb). Why does it make sense to get this answer? *Hint: take a look at the column `Ozone` in the dataset.*

```{r}
# Loading the data set airquality
data("airquality")
# Calculating the mean of Ozone (in ppb)
mean(airquality$Ozone)
```

**Calculating the mean of the Ozone and the output being NA makes sense because the data set has values listed as NA.**

------------------------------------------------------------------------

### Question 8: (2 pts)

Look at the documentation for the function `mean()` by running `?mean` **in the console**. What argument should be used to find the mean value that we were not able to get in the previous question? What type of values does that argument take?

**After looking at the documentation for `mean()` the arugement that should be used to find the mean value is na.rm. This argument takes the values `TRUE` or `FALSE`.**

------------------------------------------------------------------------

### Question 9: (2 pts)

Sometimes the R documentation does not feel complete. We wish we had more information or more examples. Find a post online (include the link) that can help you use that argument in the `mean()` function. Then finally find the mean ozone!

[R Documentation For The `mean()` Function](https://www.educative.io/answers/what-is-the-mean-method-in-r)
```{r}
# Calculating the mean of Ozone using the arguments
mean(airquality$Ozone, na.rm = TRUE)
```

**After setting the na.rm argument to `TRUE` and calculating the mean. We calculated that the mean of the variable `Ozone` is 42.12931 ppb.**

------------------------------------------------------------------------

## Part 3: (5 pts)

The Internet clothing retailer Stitch Fix has developed a new model for selling clothes to people online. Their basic approach is to send people a box of 5--6 items of clothing and allow them to try the clothes on. Customers keep (and pay for) what they like while mailing back the remaining clothes. Stitch Fix then sends customers a new box of clothes typically a month later.

A critical question for Stitch Fix to consider is "Which clothes should the send to each customer?" Since customers do not request specific clothes, Stitch Fix has to come up with 5--6 items on its own that it thinks the customers will like (and therefore buy). In order to learn something about each customer, they administer an **intake survey** when a customer first signs up for the service. The survey has about 20 questions and the data is then used to predict what kinds of clothes customers will like. In order to use the data from the intake survey, a statistical algorithm must be built in order to process the customer data and make clothing selections.

Suppose you are in charge of building the intake survey and the algorithm for choosing clothes based on the intake survey data. 


### Question 10: (2 pts)

What kinds of questions do you think might be useful to ask of a customer in an intake survey in order to better choose clothes for them? What kinds of data would be most valuable? See if you can come up with at least 5 items.

**When creating an intake survey there are a few important questions that customers should be asked:**

* **What is your height?**
* **What is your weight?**
* **What kind of colors do you like?**
* **Do you like your clothes to fit loose or tight?**
* **What stores do you like to shop at?**


### Question 11: (3 pts)

In addition to the technical challenges of collecting the data and building this algorithm, you must also consider the impact the algorithm may have on the people involved. What potential negative impact might the algorithm have on the customers who are submitting their data? Consider both the data being submitted as well as the way in which the algorithm will be used when answering this question.


**With any kind of survey or algorithm there are bound to be unintended negative consequences. The algorithm may suggest clothing that the customer may never wear which may make the customer question if their wardrobe is adequate. It may also suggest the wrong sizes for customers which could have a detrimental impact on a customer's body image.**







------------------------------------------------------------------------

## Formatting: (3 pts)

Knit your file! Into pdf directly or into html. 

Is it working? If not, try to decipher the error message (look up the error message, consult websites such as [stackoverflow](https://stackoverflow.com/) or [crossvalidated](https://stats.stackexchange.com/). 

Once it knits in html, click on `Open in Browser` at the top left of the window pops out. Print your html file into pdf from your browser. Any issue? Ask your classmates or TA!

------------------------------------------------------------------------

```{r, echo=F}
## DO NOT DELETE THIS BLOCK!
sessionInfo()
```