---
title: "HW 2"
output: 
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
editor_options: 
  chunk_output_type: console
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE, fig.align = "center", warning = F, message = F, R.options=list(max.print=100))

# Edit the file starting below
```

## Andrew Manrriquez (ajm7246)

**This homework is due on September 15, 2023 by 11:59pm. You will submit a pdf file on Gradescope.**

*For all questions, include the R code that you used to find your answer (show R chunk and outputs). Answers without supporting code will not receive credit (unless no code was required). Outputs without comments will not receive credit either: Write full sentences to describe your findings.*

------------------------------------------------------------------------

### Question 1: (2 pts)

##### The dataset `ChickWeight` contains information about the weights (in grams) of chicks on four different diets over time (measured at 2-day intervals) as the result of an experiment. The first few observations are listed below.

```{r}
# Save dataset as a dataframe
ChickWeight <- as.data.frame(ChickWeight)

# Visuaize the first ten rows of the dataset
head(ChickWeight, 10)
```

Use some combination of `table()` and `length()` to answer the following questions: 

- How many distinct chicks are there?  
- How many distinct time points?  
- How many distinct diet conditions?  
- How many chicks per diet condition?  

```{r}
# Distinct chicks?
length(unique(ChickWeight$Chick))
# Distinct time points?
length(unique(ChickWeight$Time))
# Distinct diet conditions?
length(unique(ChickWeight$Diet))
# How many chicks per diet?
table(ChickWeight$Diet)
```

**There are 50 distinct chicks, 12 distinct time points (days), 4 distinct diets. For diet 1 there are 220 chicks, diet 2 are 120 chicks, diet 3 are 120 chicks, and for diet 4 there are 118 chicks.**

------------------------------------------------------------------------

### Question 2: (2 pts)

##### Using the `ggplot2` package, create a simple scatterplot showing chick `weight` (on the y-axis) as a function of `Time`. Label the axes including the units of the variables and give the plot a title. How does chick `weight` change over `Time`?

```{r}
# Load the tidyverse (including ggplot2)
library(tidyverse)
library(ggplot2)

# Creating a scatterplot
ChickWeight %>% ggplot(aes(x = Time, y = weight)) + 
        geom_point(color = "#5579C6") + 
        labs(title = "Chick Weight Over Time", x = "Time (Days Since Birth)", y = "Weight (gm)") + 
        theme_minimal()
```

**The weight of a chick increases as time increases, indicative of a positive relationship between the two variables.**

------------------------------------------------------------------------

### Question 3: (2 pts)

##### Building upon the previous plot, map `Chick` to an aesthetic that assigns a color to each chick’s data points. Add lines that connect each chick’s points together with `geom_line()`. Finally, remove the legend. Do all chicks seem to gain weight in the same manner? Why/Why not?

```{r}
# Reordered Chicks in numerical order
ChickWeight <- ChickWeight %>% 
        mutate(Chick = factor(Chick, 
                             levels = c("1", "2", "3", "4", "5", 
                                        "6", "7", "8", "9", "10", 
                                        "11", "12", "13", "14", "15", "16", "17", "18", "19",
                                        "20", "21","22","23","24", "25", "26", "27",
                                        "28","29","30","31","32","33","34","35","36","37",
                                        "38","39","40","41","42","43","44","45","46","47","48","49","50")))

# Creating a scatter plot with different colors for each chick
ChickWeight %>% ggplot(aes(x = Time, y = weight)) + 
        geom_point(aes(color = Chick)) + 
        geom_line(aes(color = Chick, linetype = Chick)) + 
        labs(title = "Chick Weight Over Time", x = "Time (Days Since Birth)", y = "Weight (gm)") +
        theme(legend.position = "none")
 
```

**Every chick seems to gain weight in a different manner. Reasoning's behind this might be that the chicks might have different metabolisms, eat more or less than other chicks, or are more/less active.**

------------------------------------------------------------------------

### Question 4: (2 pts)

Continue modifying the same plot by 

* removing the color from the points only
* make all of the points white 
* leave the lines colored by chick
* Put the points *on top of* the lines 

On which day was the last value of the chicks' weight recorded?

```{r}
# Modifying the plot
ChickWeight %>% ggplot(aes(x = Time, y = weight)) + 
        geom_line(aes(color = Chick, linetype = Chick)) + 
        geom_point(color = "white", size = 2.5) + 
        labs(title = "Chick Weight Over Time", x = "Time (Days Since Birth)", y = "Weight (gm)") +
        theme_dark() + 
        theme(legend.position = "none") + 
        scale_x_continuous(breaks = seq(0,25,1))

```

**On the 21st day the last value of the chicks' weight was recorded.**

------------------------------------------------------------------------

### Question 5: (2 pts)

Now, facet this plot by diet. Can you tell from this new plot which diet results in greater weight? Explain.

```{r}
# Facet wrapping the plot by diet
ChickWeight %>% ggplot(aes(x = Time, y = weight)) + 
        geom_line(aes(color = Chick, linetype = Chick)) + 
        geom_point(color = "white", size = 2) + 
        facet_wrap(vars(Diet)) +
        labs(title = "Chick Weight Over Time", x = "Time (Days Since Birth)", y = "Weight (gm)") +
        theme_dark() + 
        theme(legend.position = "none") + 
        scale_x_continuous(breaks = seq(0,25,1))
```

**From the facet plots it is hard to tell which diet results in greater weight. This is because `geom_line` only creates 13 different lines and we have 50 chicks to keep track of.**

------------------------------------------------------------------------

### Question 6: (2 pts)

Go back to your plot from question 2 and fit a *linear regression line* (using `lm`) to the chicks in each diet with `geom_smooth()`. There should be 4 separate regression lines, one for each diet, each a separate color. Can you see more clearly which diet results in greater weight? Explain.

```{r}
# Creating a scatterplot
ChickWeight %>% ggplot(aes(x = Time, y = weight)) + 
        geom_point(color = "#5579C6") + 
        geom_smooth(aes(color = Diet)) +
        labs(title = "Chick Weight Over Time", x = "Time (Days Since Birth)", y = "Weight (gm)") + 
        theme_minimal()
```

**Using `geom_smooth` it makes it more clear to see which diet results in a greater weight. It is easier to see because each diet is a different color so it is easy to distinguish.**

------------------------------------------------------------------------

### Question 7: (2 pts)

A scatterplot might not be the best way to visualize this data: it calls attention to the relationship between weight and time, but it can be hard to see the differences between diets. A more traditional approach for exploring the effect of diet would be to construct a barplot representing group means with standard error bars showing +/- 1 standard error.

Create a plot using `geom_bar()` where each bar's height corresponds to the average chick weight for each of the four diet conditions. Rename the y-axis to include units (e.g., with scale_y_continuous(`name=`...)) and make the major tick marks go from 0 to 150 by 25 (e.g., with scale_y_continuous(`breaks=`...)). Which diet has the highest mean `weight`? 

```{r}
# Create a bar plot 
ChickWeight %>% ggplot(aes(x = Diet, y = weight, fill = Diet)) + 
        geom_bar(stat = "summary", fun = "mean", alpha = 0.8) +
        scale_y_continuous(name = "Average Chick Weight (gm)", breaks = seq(0,150,25)) + 
        labs(title = "Average Weight of Chicks Based on Diets", x = "Diet", y = "Average Chick Weight (gm)")
```

**Based on the bar plot diet 3 has the highest mean weight.**

------------------------------------------------------------------------

### Question 8: (2 pts)

Add error bars showing + or - 1 SE using `geom_errorbar(stat = "summary")`. Make the error-bars skinnier by adding a `width =`**0.5** argument. Color the bars (not the error bars, but the barplot bars) by diet and change from the default color scheme using a `scale_fill_` or a `scale_color_`. diet seems to have the most variation in `weight`? The least variation?

```{r}
library(RColorBrewer)
# Creating a bar plot with error bars
ChickWeight %>% ggplot(aes(x = Diet, y = weight, fill = Diet)) + 
        geom_bar(stat = "summary", fun = "mean", alpha = 0.8) +
        scale_fill_brewer(type = "seq", palette = "RdBu", direction = 1) + 
        geom_errorbar(stat = "summary", width = 0.5, fun.data = "mean_se") +
        scale_y_continuous(name = "Average Chick Weight (gm)", breaks = seq(0,150,25)) + 
        theme_minimal() + 
        labs(title = "Barplot with Error Bars By Diet", x = "Diet", y = "Average Chick Weight (gm)") 
```

**Diet 3 has the most variation in weight and diet 1 has the least variation in weight.**

------------------------------------------------------------------------

### Question 9: (2 pts)

Take your code from question 8 and replace `geom_bar()` with `geom_point()`. Remove the `breaks=` argument from `scale_y_continuous`. Make the points larger and color them all red. Put them *on top of* the error bars. Does the mean chick weight seem to differ based on the diet? *I am not asking to conduct hypothesis testing but informally state if they seem to differ and if so, how.*

```{r}
## add your code here (and edit this comment)
ChickWeight %>% ggplot(aes(x = Diet, y = weight)) + 
        geom_errorbar(stat = "summary", width = 0.5, fun.data = "mean_se") +
        geom_point(stat = "summary", fun = "mean", size = 10, aes(color = "red")) +
        theme_minimal() +
        scale_fill_brewer(type = "seq", palette = "RdBu", direction = 1) + 
        theme(legend.position = "none") + 
        scale_y_continuous(name = "Average Chick Weight (gm)") + 
        labs(title = "Average Weight of Chicks Based on Diets", x = "Diet", y = "Average Chick Weight (gm)")
```

**When using `geom_point` instead of `geom_bar` there is still a difference in the mean chick weight for each diet. Diet 3 still has the highest mean weight and Diet 1 still has the lowest mean weight.**

------------------------------------------------------------------------

### Formatting: (2 pts)

##### Comment your code, write full sentences, and knit your file!

------------------------------------------------------------------------