---
title: "HW 4"
output: html_document
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE,  
                      warning = FALSE, message = FALSE, 
                      fig.align = "center",
                      R.options = list(max.print=100))

# Edit the file starting below
```

### Andrew Manrriquez

**You will submit this homework assignment as a pdf file on Gradescope.**

*For all questions, include the R commands/functions that you used to find your answer (show R chunk). Answers without supporting code will not receive credit. Write full sentences to describe your findings.*

------------------------------------------------------------------------

### Question 1: (2 pts)

All subsequent code will be done using `dplyr`, so we need to load this package. We also want to look at the `penguins` dataset which is inside the `palmerpenguins` package:

```{r message=FALSE}
# Call dplyr and ggplot2 packages within tidyverse
library(tidyverse)

# Paste and run the following uncommented code into your console: 
# install.packages("palmerpenguins")

# Save the data as a dataframe
penguins <- as.data.frame(palmerpenguins::penguins)
```

Using a `dplyr` function, pick all the rows/observaions in the `penguins` dataset from the year 2007 and save the result as a new object called `penguins_2007`. Compare the number of observations/rows in the original `penguins` dataset with your new `penguins_2007` dataset.

```{r}
# Creating an object called "penguin_2007"
penguins_2007 <- penguins %>% filter(year == 2007)
# Number of rows in original data set
nrow(penguins)
# Number of rows in 2007 penguin data set
nrow(penguins_2007)
```

**There are 344 observations/rows in the original penguins data set and there are 110 observations/rows in the new `penguins_2007` dataset.**

------------------------------------------------------------------------

### Question 2: (2 pts)

Using `dplyr` functions on `penguins_2007`, report the number of observations for each species-island combination (note that you'll need to `group_by`). Which species appears on all three islands? 

```{r}
# Grouping by island and species
species_island_count <- penguins_2007 %>%
        group_by(species, island) %>% 
        summarise(count = n())
# Printing the count for each species-island combo
print(species_island_count)
# Finding species that are on all three islands
species_all_islands <- species_island_count %>% 
        group_by(species) %>% 
        filter(n_distinct(island) == 3) %>% 
        select(species) %>% 
        distinct()
print(species_all_islands)
        

```

**There are 10 penguins of the Adelie species and 34 Gentoo on the Biscoe island. On the Dream island there are 20 penguins of the Adelie species and 26 penguins of the Chinstrap species. On the Torgersen island there are 20 penguins of the Adelie species. Only the Adelie species appears on all three islands.**

------------------------------------------------------------------------

### Question 3: (2 pts)

Using `dplyr` functions on `penguins_2007`, create a new variable that contains the ratio of `bill_length_mm` to `bill_depth_mm` (call it `bill_ratio`). *Once you checked that your variable is created correctly*, overwrite `penguins_2007` so it contains this new variable.

```{r}
# Creating variable for ratio of bill length and depth
penguins_2007 <- penguins_2007 %>%
        mutate(bill_ratio = bill_length_mm / bill_depth_mm)
```

Are there any cases in the `penguins_2007` dataset for which the `bill_ratio` exceeds 3.5? If so, for which species of penguins is this true?

```{r}
# Finding which species has bill ratio greater than 3.5 
penguins_2007 %>% 
        filter(bill_ratio > 3.5) %>% 
        select(species) %>%
        distinct()
```

**The Gentoo species has a bill ratio greater than 3.5.**

------------------------------------------------------------------------

### Question 4: (2 pts)

Using `dplyr` functions on `penguins_2007`, find the three penguins with the smallest bill ratio for *each species*. Only display the information about `species`, `sex`, and `bill_ratio`. Does the same sex has the smallest bill ratio across species?

```{r}
# Finding three penguins with smallest bill ratio for each speices
top_small_ratio <- penguins_2007 %>%
        arrange(species, bill_ratio) %>%
        group_by(species) %>%
        slice_head(n = 3) %>%
        select(species, sex, bill_ratio)
print(top_small_ratio)

```

**For the Adelie species the three smallest bill ratios are from males and for the Chinstrap species the three smallest bill ratios are from females. For the Gentoo species the three smallest bill ratios are from two females and a male.**

------------------------------------------------------------------------

### Question 5: (2 pts)

Using `dplyr` functions on `penguins_2007`, calculate the mean and standard deviation of `bill_ratio` for each species. Drop NAs from `bill_ratio` for these computations (e.g., using the argument `na.rm = T`) so you have values for each species. Which species has the greatest mean `bill_ratio`?

```{r}
# Calculating mean and standard deviation of bill ratio
species_bill_ratio_summary <- penguins_2007 %>%
        group_by(species) %>%
        summarise(mean_bill_ratio = mean(bill_ratio, na.rm = T), sd_bill_ratio = sd(bill_ratio, na.rm = T))
print(species_bill_ratio_summary)

greatest_mean_bill_ratio <- species_bill_ratio_summary %>%
        filter(mean_bill_ratio == max(mean_bill_ratio)) %>%
        select(species, mean_bill_ratio)
print(greatest_mean_bill_ratio)

```

**The Gentoo species has the greatest mean bill ratio of 3.20.**

------------------------------------------------------------------------

### Question 6: (2 pts)

Using `dplyr` functions on `penguins_2007`, remove missing values for `sex`. Pipe a `ggplot` to create a single plot showing the distribution of `body_mass_g` colored by male and female penguins, faceted by species (use the function `facet_wrap()` with the option `nrow = ` to give each species its own row). Which species shows the least sexual dimorphism (i.e., the greatest overlap of male/female size distributions)?

```{r}
# Removing missing values for sex
penguins_2007_filtered <- penguins_2007 %>% filter(!is.na(sex))

# Creating a plot
penguins_2007_filtered %>% ggplot(aes(x = body_mass_g, fill = sex)) + 
        geom_density(alpha = 0.8) + 
        facet_wrap(~species, nrow = 3) + 
        labs(title = "Distribution of Body Mass by Sex and Species", x = "Body Mass (g)", y = "Density") + theme_minimal()

```

**The Chinstrap species shows the least sexual dimorphism.**

------------------------------------------------------------------------

### Question 7: (2 pts)

Pipe`penguins_2007` to `ggplot()` to create a scatterplot of `body_mass_g` (y-axis) against `bill_ratio` (x-axis). Does it look like there is a relationship between the bill ratio and the body mass? *Note: you might see a Warning message.* What does this message refer to?*

```{r}
# Creating a scatterplot
penguins_2007 %>% ggplot(aes(x = bill_ratio, y = body_mass_g)) + 
        geom_point() + 
        labs(x = "Bill Ratio", y = "Body Mass (g)", title = "Scatterplot of Body Mass vs Bill Ratio")

```

**There appears to be a positive relationship between bill ratio and body mass. As the bill ratio increases the body mass increases as well. The error message refers to a row in the data set having missing values.**

------------------------------------------------------------------------

### Question 8: (2 pts)

What if we separate each species? Duplicate the plot from the previous question and add a regression trend line with `geom_smooth(method = "lm")`. Color the points AND the regression lines by species. Does the relationship between the bill ratio and the body mass appear to be the same across the different species?

```{r}
# Creating a scatterplot
penguins_2007 %>% ggplot(aes(x = bill_ratio, y = body_mass_g, color = species)) + 
        geom_point() + 
        geom_smooth(method = "lm") + 
        facet_wrap(vars(species))
        labs(x = "Bill Ratio", y = "Body Mass (g)", title = "Scatterplot of Body Mass vs Bill Ratio") + 
        theme_minimal()
```

**The relationship between the bill ratio and body mass is similar for the Gentoo and Chinstrap species, with both having a positive relationship. However for the Adelie species it has a negative relationship.**

------------------------------------------------------------------------

### Question 9: (2 pts)

Finally, let's make a plot using the original `penguins` dataset (not just the 2007 data). Forewarning: This will be very busy plot!

Map `body_mass_g` to the y-axis, `flipper_length_mm` to the x-axis, `species` to color, and `island` to shape. Using `facet_wrap()`, facet the plots by `year`. Find a way to clean up the x-axis labels (e.g., reduce the number of tick marks) using `scale_x_continuous()`. Does there appear to be a relationship between body mass and flipper length overall? Is there a relationship within each species? What happens to the distribution of flipper lengths for species over time?

```{r}
# Creating a plot with original data set
penguins %>% ggplot(aes(x = flipper_length_mm, y = body_mass_g, color = species, shape = island)) + 
        geom_point() + 
        facet_wrap(vars(year)) + 
        labs(x = "Flipper length (mm)", y = "Body Mass (g)", title = "Scatterplot of Body Mass vs Flipper Length") + 
        scale_x_continuous(breaks = seq(170,230,25)) +
        theme_minimal()

```

**There appears to be a positive relationship between body mass and flipper length; as flipper length increases body mass increases. Within each species the relationship between flipper length and body mass appears to be positive as well. The distribution of flipper length for each species over time appears to shift towards larger flipper lengths.**

------------------------------------------------------------------------

### Formatting: (2 pts)

Comment your code, write full sentences, and knit your file!

