---
title: "HW 5"
output: html_document
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE,  
                      warning = FALSE, message = FALSE, 
                      fig.align = "center",
                      R.options = list(max.print=100))

# Edit the file starting below
```

### Andrew Manrriquez (ajm7246)

**You will submit this homework assignment as a pdf file on Gradescope.**

*For all questions, include the R commands/functions that you used to find your answer (show R chunk). Answers without supporting code will not receive credit. Write full sentences to describe your findings.*

**NOTE**: You must use version 1.3.0 of the `tidyr` package for this homework. If you are not sure what version you have, you can run `install.packages("tidyr")` in the console window and R will install the latest version of the package.

------------------------------------------------------------------------

### Question 1: (1 pt)

The dataset `world_bank_pop` is a built-in dataset in `tidyverse`. It contains information about total population and population growth, overall and more specifically in urban areas, for countries around the world. Take a look at it with `head()`. Is the data tidy? Why or why not?

```{r message=FALSE}
# Call tidyr, dplyr and ggplot2 packages within tidyverse
library(tidyverse)

world_bank_pop

# Take a look!
head(world_bank_pop)
print(world_bank_pop)
```

**The `world_pop` data set is not tidy. This data set is not tidy because the variables are stored in rows as opposed to in columns.**

------------------------------------------------------------------------

### Question 2: (1 pt)

Using `dplyr` functions on `world_bank_pop`, count how many distinct countries there are in the dataset. Does this makes sense? Why or why not?

```{r}
# Load necessary libraries and assume 'world_bank_pop' is a dataframe with a 'country' column
library(dplyr)

# Assuming 'world_bank_pop' is the correct dataframe name and 'country' is a column
distinct_countries <- world_bank_pop %>% 
                      distinct(country) %>% 
                      count()

# Print the count of distinct countries
print(distinct_countries)

```

**There are 266 distinct countries in the data set. This does not make sense because there are only 195 countries in the world.**

------------------------------------------------------------------------

### Question 3: (2 pts)

Use one of the `pivot` functions on `world_bank_pop` to create a new dataset with the years 2000 to 2017 appearing as a *numeric* variable `year`, and the different values for the indicator variable are in a variable called `value`. Save this new dataset in your environment as `myworld1`. 

```{r}
# Pivot the dataset
myworld1 <- world_bank_pop %>%
  pivot_longer(cols = starts_with("20"), 
               names_to = "year", 
               values_to = "value")

# Print the first few rows of the resulting dataset
head(myworld1)

```

How many rows are there per country? Why does it make sense?

```{r}
# Grouping by country
rows_per_country <- myworld1 %>%
        group_by(country) %>%
        count()
print(rows_per_country)

```

**There are 72 rows per country.This makes sense because there are four different indicators and there are 18 time points of data, thus this accounts for the 72 rows of data per country.**

------------------------------------------------------------------------

### Question 4: (3 pts)

Use another `pivot` function on `myworld1` to create a new dataset, `myworld2`, with the different categories for the indicator variable appearing as their own variables. Use `dplyr` functions to rename `SP.POP.GROW` and `SP.URB.GROW`, as `pop_growth` and `pop_urb_growth` respectively. 

```{r}
# Pivot data to create new variables
myworld2 <- myworld1 %>%
        pivot_wider(names_from = indicator, values_from = value)

# Renaming columns
myworld2 <- myworld2 %>% 
        rename(pop_growth = "SP.POP.GROW",
               pop_urb_growth = "SP.URB.GROW")

head(myworld2)
```

Using `dplyr` functions, find the ratio of urban growth compared to the population growth in the world for each year. *Hint: the country code `WLD` represents the entire world.* Create a `ggplot` to display how the percentage of urban population growth has changed over the years. Why does your graph not contradict the fact that the urban population worldwide is increasing over the years?

```{r}
# Filter for the world (country code 'WLD') and aggregate the ratio by year
world_ratio <- myworld2 %>%
  filter(country == "WLD") %>%
  group_by(year) %>%
  summarise(urban_to_population_ratio = mean(pop_urb_growth / pop_growth) * 100)

# Print the aggregated data to inspect
print(world_ratio)

# Create a ggplot to display the percentage of urban population growth over the years
ggplot(world_ratio, aes(x = year, y = urban_to_population_ratio)) +
  geom_point() +
  labs(x = "Year",
       y = "Urban Population Growth (%)",
       title = "Percentage of Urban Population Growth Over the Years")

```

**The graph does not contradict the idea that the urban population worldwide is increasing over the years because we are plotting the urban population growth by percentage over years. This percentage can still increase even while urban population and total population is increasing.**

------------------------------------------------------------------------

### Question 5: (1 pt)

In `myworld2`, which country code had the highest population growth in 2017? *Hint: Use the `arrange()` function here.*

```{r}
# Filter for 2017 and arrange
highest_growth_2017 <- myworld2 %>%
  filter(year == 2017) %>%
  arrange(desc(pop_growth)) %>%
  slice(1)

# Print the country code with the highest population growth in 2017
print(highest_growth_2017$country)


```

**QAT had the highest population growth in 2017.**

------------------------------------------------------------------------

### Question 6: (1 pt)

When answering the previous, we only reported the three-letter code and (probably) have no idea what the actual country is. We will now use the package `countrycode` with a built-in dataset called `codelist` that has information about the coding system used by the World bank:

Using `dplyr` functions, modify `mycodes` above to only keep the variables `continent`, `wb` (World Bank code), and `country.name.en` (country name in English). Then remove countries with missing `wb` code.

```{r}
# Paste and run the following into your console (NOT HERE): install.packages("countrycode")

# Call the countrycode package
library(countrycode)

# Create a list of codes with matching country names
mycodes <- codelist %>%
        select(continent, wb, country.name.en) %>%
        filter(!is.na(wb))

```

How many countries are there in `mycodes`? 

```{r}
# Count number of countries
num_countries <- nrow(mycodes)
print(num_countries)
```

**There are 218 countries in `mycodes`.**

------------------------------------------------------------------------

### Question 7: (1 pt)

Use a `left_join()` function to add the information of the country codes **to** `myworld2` dataset. Match the two datasets based on the World Bank code. *Note: the column containing the World Bank code does not have the same name in each dataset.* Using `dplyr` functions, only keep the data available for Europe and for the year 2017. Save this new dataset as `myeurope`.

```{r}
# your code goes below (make sure to edit comment)
myeurope <- myworld2 %>%
        filter(year == 2017) %>%
        left_join(mycodes, by = c("country" = "wb")) %>%
        filter(continent == "Europe")
```

How many rows are there in this new dataset `myeurope`? What does each row represent?

```{r}
# Counting the rows
nrow(myeurope)
```

**There are 46 rows in the `myeurope` data set. Each row represents a country that is in Europe.**

------------------------------------------------------------------------

### Question 8: (2 pts)

Using `dplyr` functions on `myeurope`, only keep information for the population growth in 2017 then compare the population growth per country with `ggplot` using `geom_bar()`. Use the `reorder()` function to order countries in order of population growth. Which country in Europe had the lowest population growth in 2017?

```{r}
# 2017 and population growth filter
population_growth_2017 <- myeurope %>%
  filter(year == 2017) %>%
  group_by(country.name.en) %>%
  summarise(pop_growth = sum(pop_growth, na.rm = TRUE))

# Order countries by population growth
population_growth_2017 <- population_growth_2017 %>%
  arrange(reorder(country.name.en, -pop_growth))

# Create a ggplot to visualize population growth per country
ggplot(population_growth_2017, aes(x = reorder(country.name.en, -pop_growth), y = pop_growth)) +
  geom_bar(stat = "identity", fill = "skyblue") +
  labs(x = "Country", y = "Population Growth", title = "Population Growth per Country in Europe (2017)") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Find the country with the lowest population growth
lowest_growth_country <- population_growth_2017 %>% 
  filter(pop_growth == min(pop_growth, na.rm = TRUE))
print(lowest_growth_country)

 
```

**Moldova had the lowest population growth in 2017 at -1.727058%.**

------------------------------------------------------------------------

### Question 9: (1 pt)

When dealing with location data, we can actually visualize information on a map if we have geographic information such as latitude and longitude. Next, we will use a built-in function called `map_data()` to get geographic coordinates about countries in the world (see below). Take a look at the dataset `mapWorld`. What variables could we use to join `mapWorld` and `myeurope`? *Note: the variables do not have the same name in each dataset but they contain the same information.*

```{r}
# Geographic coordinates about countries in the world
mapWorld <- map_data("world") %>% 
        as_tibble()

# Rename the variable in 'mapWorld' to match the variable name in 'myeurope'
mapWorld <- mapWorld %>%
  rename(country.name.en = region)

# Join the data sets based on the country names
merged_data <- myeurope %>%
  inner_join(mapWorld, by = "country.name.en")

# Print the data set
head(merged_data)

```

**We could could use the `country.name.en` and `region` variables in both data sets to join the two.**

------------------------------------------------------------------------

### Question 10: (2 pts)

Use a joining function to check if any information from `myeurope` is not contained in `mapWorld`, matching the two datasets based on the country name.

```{r}
# Perform a left join
merged_data <- myeurope %>%
  left_join(mapWorld, by = "country.name.en")

# Filter for rows where there is no match in 'mapWorld'
missing_info <- merged_data %>%
  filter(is.na(long))

# Print the countries from 'myeurope' with missing information in 'mapWorld'
print("Countries from 'myeurope' with missing information in 'mapWorld':")
print(missing_info)

```

Some countries such as United Kingdom did not have a match. Why do you think this happened? *Hint: find the distinct country names in `mapWorld`, arrange them in alphabetical order, and scroll through the names. Can you find any of these countries with no match in a slightly different form?* If you need to print more output from a tibble, you can use `print(n = X)` where `X` is the number of lines to print out.

```{r}
# Print distinct country names in 'mapWorld' in alphabetical order
distinct_countries_mapWorld <- mapWorld %>%
  distinct(country.name.en) %>%
  arrange(country.name.en)

print(distinct_countries_mapWorld, n = Inf)

```

**These countries may not have a match due to the variations in their country name. The variations may lead to issues with finding exact matches in both data sets.**

------------------------------------------------------------------------

### Question 11: (1 pt)

Consider the `myeurope` dataset. Recode some of the country names so that the countries with no match from the previous question (with the exception of Gibraltar which is not technically a country anyway) will have a match. 

*Hint: use `recode()` inside `mutate()` as described in this article https://www.statology.org/recode-dplyr/.* Then add a pipe and use a `left_join()` function to add the geographic information in `mapWorld` to the countries in `myeurope`. Save this new dataset as `mymap`.

```{r}
# Recode country names to match those in 'mapWorld'
recoded_myeurope <- myeurope %>%
  mutate(country.name.en = recode(country.name.en,
                                  "United Kingdom" = "Great Britain",
                                  "Bosnia & Herzegovina" = "Bosnia and Herzegovina", 
                                  "Czechia" = "Czech Republic"
                                  )) 

# Perform a left join with 'mapWorld' based on the recoded country names
mymap <- recoded_myeurope %>%
  left_join(mapWorld, by = "country.name.en")

# Print the first few rows of the resulting data set
head(mymap)

```

------------------------------------------------------------------------

### Question 12: (2 pts)

Let's visualize how population growth varies across European countries in 2017 with a map. Use the R code provided below. Add a comment after each `#` to explain what each component of this code does. *Note: it would be a good idea to run the code piece by piece to see what each layer adds to the plot.*

```{r}
# Build a map!
mymap %>%
  # Creating a base plot using ggplot
  ggplot(aes(x = long, y = lat, group = group, fill = pop_growth)) + 
  # This adds polygons to the plot based off the longitude and latitude columns
  geom_polygon(colour = "black") +
  # This will color the map on a scale with white being the low and blue the highest
  scale_fill_gradient(low = "white", high = "blue") +
  # Adding labels to the plot
  labs(fill = "Growth" ,title = "Population Growth in 2017", 
       x ="Longitude", y ="Latitude") +
  # Setting the limit for the longitude (x-axis) and latitude (y-axis)
  xlim(-25,50) + ylim(35,70)

```

Which country had the highest population growth in Europe in 2017? *Hint: it's very tiny! You can refer to this map for European geography: https://www.wpmap.org/europe-map-hd-with-countries/*

```{r}
# Filter for the year 2017 and identify the country with the highest population growth
highest_growth_country <- mymap %>%
  filter(year == 2017) %>%
  arrange(desc(pop_growth)) %>%
  slice(1)  # Select the top row (highest population growth)

# Print the country with the highest population growth in 2017
print(highest_growth_country)

```


**Malta had the highest population growth in Europe in 2017 at 2.738663%.**

------------------------------------------------------------------------

### Formatting: (2 pts)

Comment your code, write full sentences, and knit your file!

------------------------------------------------------------------------

```{r, echo=F}
## DO NOT DELETE THIS BLOCK!
Sys.info()
```