---
title: "HW 6"
output: html_document
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE,  
                      warning = FALSE, message = FALSE, 
                      fig.align = "center",
                      R.options = list(max.print=100))

# Edit the file starting below
```

### Andrew Manrriquez (ajm7246)

**You will submit this homework assignment as a pdf file on Gradescope.**

*For all questions, include the R commands/functions that you used to find your answer (show R chunk). Answers without supporting code will not receive credit. Write full sentences to describe your findings.*

We will use the packages `tidyverse`, `factoextra`, and `cluster` for this assignment.

```{r, message = FALSE}
# Load packages
library(tidyverse)
library(factoextra)
library(cluster)
```

------------------------------------------------------------------------

## Question 1: (2 pts)

The dataset for this homework comes from the article:

*Tsuzuku N, Kohno N. 2020. The oldest record of the Steller sea lion Eumetopias jubatus (Schreber, 1776) from the early Pleistocene of the North Pacific. <https://doi.org/10.7717/peerj.9709>*

Read the ***Abstract*** of the article and the section called *Results of Morphometric Analyses*. What was the goal of this study and what was the main finding?

**The goal of this study, was to investigate and gain insight into the early Pleistocene history of the Steller sea lion andEuntopias jubatus in the North Pacific region. The main finding was the discovery of the oldest record of the Steller sea lion.**

------------------------------------------------------------------------

## Question 2: (1 pt)

Under the supplemental information, I retrieved the data from a word document into a `.csv` document. Import the dataset from GitHub. 

```{r}
# download data from GitHub
sealions <- read_csv("https://raw.githubusercontent.com/laylaguyot/datasets/main//Sealions.csv")
```

How many rows and how many columns are in this dataset? What does a row represent? What does a column represent?

```{r}
# your code goes below (make sure to edit comment)
nrow(sealions)
ncol(sealions)
```

**There are 51 rows and 39 columns in this data set. Each row represents an individual sea lion and each column represents a specific feature/attribute of sea lions that were measured.**

------------------------------------------------------------------------

## Question 3: (1 pt)

Before we can analyze the data, let's do some cleaning. Using a combination of the `select()`, `where()`, and a predicate function like `is.character()` we can scan through all the columns of the dataset and see which columns are of character type.

```{r}
## Select all of the columns that have character type
sealions %>% 
    select(where(is.character))
```

When importing this dataset into RStudio, which variables were considered numeric? Why are some measurements not considered as numeric? Use the `is.numeric()` predicate function here.


```{r}
# your code goes below (make sure to edit comment)
sealions %>% 
        select(where(is.numeric))
```

**The variables `K` and `AD` were considered numeric. Some measurements may not be considered numeric due to formatting or non-standard representations in the original data set. Examples of this may include non-numeric characters being used, missing data, and inconsistent formatting.**

------------------------------------------------------------------------

## Question 4: (1 pt)

The functions `mutate()` and `across()`, when used together, can make changes across a range of columns in a data frame. 

Using `mutate()` and `across()`, replace all `-` in the dataset with the missing values *NA* and then then make sure all measurements are defined as numeric variables with. The first part of the code replaces the "-" with NA values. Write the second part to coerce all of the columns (except for the ID column!) to be numeric. Make sure to overwrite the dataset `sealions`. 

NOTE: Look at the examples in `?across` to get a sense of how to use the `across()` function.

```{r}
# Mutating the original data set
sealions <- sealions %>% 
    ## Replace all "-" with NA
    mutate(across(where(is.character), ~ na_if(.x, "-"))) %>%
        ## Coerce all columns into numeric except `ID`
        mutate(across(-ID, as.numeric))
```

What is the mean rostral tip of mandible `C`?

```{r}
# Calculating the mean
mean_rostral_tip <- mean(sealions$C, na.rm = TRUE)
# Printing the mean
print(mean_rostral_tip)

```

**The mean rostral tip of mandible is 34.86622 mm.**

------------------------------------------------------------------------

## Question 5: (2 pts)

You are given the code in this question. But what does the code do? Write comments.

```{r}
sealions <- sealions %>% 
  ## 
  select_if(!is.na(sealions[51,])) %>% 
  ## 
  na.omit

nrow(sealions)
ncol(sealions)
```

How many columns and how many rows are remaining in this dataset?

**The code first selects columns where the 51st row does not contain missing values. It then removes rows with missing values (NA). There are 42 rows and 23 columns in this data set.**

------------------------------------------------------------------------

## Question 6: (2 pts)

Use `dplyr` functions on `sealions` to split the `ID` variable into two variables `species` and `sex` with the function `separate()`. *Hint: in the `ID` variable, what symbol separates the species from sex?* The article states that the fossil specimen has to be male. Replace the missing value of `sex` for the fossil specimen GKZ-N 00001. *Hint: You could use the functions `mutate()` and `replace_na()`.* Save the resulting dataset as `sealions_clean`.

```{r}
# Split ID variable
sealions_clean <- sealions %>%
        separate(ID, into = c("species", "sex"), sep = " \\[|\\]", remove = FALSE)

sealions_clean <- sealions_clean %>%
        mutate(sex = replace_na(sex, "m"))
 
```

How many sealions are male/female?

```{r}
# Grouping by sex
sealions_gender <- sealions_clean %>%
        group_by(sex) %>%
        summarise(count = n())

sealions_gender

```

**There are 23 females and 19 males (18 males if we do not count the fossil).**

------------------------------------------------------------------------

## Question 7: (1 pt)

Using `dplyr` functions, only keep numeric variables and scale each numeric variable. Save the resulting dataset as `sealions_num`. What should the mean of the scaled variable of the rostral tip of mandible C be?

```{r}
# your code goes below (make sure to edit comment)
sealions_num <- sealions_clean %>%
        select_if(is.numeric) %>%
        scale()

mean_scaled_rostral_tip <- mean(sealions_num[, "C"])
print(mean_scaled_rostral_tip)
```

**The mean of the scaled variable of the rostral tip of mandible should be -9.410462e-16.**

------------------------------------------------------------------------

## Question 8: (2 pts)

Let's  perform PCA on the measurements available for the fossil specimen GKZ-N 00001. Using the function `prcomp()`, calculate the principal components (PCs) for the scaled data, `sealions_num`, obtained in the previous question. Construct a scree plot with the function `fviz_eig()` from the package `factoextra`. What is the cumulative percentage of explained variance for PC1 and PC2? 

```{r}
# Load `factoextra`
library(factoextra)

# Calculate principal components using prcomp
pca_result <- prcomp(sealions_num, scale = TRUE)

# Construct a scree plot
scree_plot <- fviz_eig(pca_result)

# Print scree plot
print(scree_plot)

# Get cumulative percentage of explained variance for PC1 and PC2 
cumulative_exp_var <- cumsum(pca_result$sdev^2)/(sum(pca_result$sdev^2))

# Extract percentage
print(cum_var_pc1 <- cumulative_exp_var[1])
print(cum_var_pc2 <- cumulative_exp_var[2])
```

**The cumulative percentageof explained variance for PC1 is 0.8476324% and for PC2 it is 0.8946766%.**

------------------------------------------------------------------------

## Question 9: (2 pts)

How many *known species* are there in `sealions_clean`? Therefore, how many clusters should we look for to identify what species GKZ-N 00001 most likely belongs to? 

```{r}
# Find number of known species
num_known_species <- length(unique(sealions_clean$species))
print(num_known_species)
```

**There are 3 known species in `sealions_clean`, this is because the fossil does not count as a species so we would not include that. Therefore we should use 3 clusters.**

Try using the PAM clustering algorithm using the `pam()` function in R (it is similar to K-means). Perform the PAM clustering algorithm on `sealions_num`, run the PAM clustering algorithm.

```{r}
# Load cluster
library(cluster)

# Perform PAM clustering
pam_results <- pam(sealions_num, k = 3)

# Print result
pam_results
```

------------------------------------------------------------------------

## Question 10: (2 pts)

Represent the clusters along the first two principal components and specify to shape the observations by their `species` in the aesthetics. *Note: you can either use `ggplot` or `fviz_cluster()`.* 

```{r}
# Load the factoextra package
library(factoextra)

# Create a scatter plot of the clusters with shape by species
cluster_plot <- fviz_cluster(pam_results, geom = "point", ggtheme = theme_minimal(), addEllipses = TRUE)

# Print the cluster plot
cluster_plot 

```

The fossil specimen GKZ-N 00001 appears to be close to which species? 

**The fossil specimen appears to be close to the Z. japonicus species.**

------------------------------------------------------------------------

## Question 11: (2 pts)

Putting it all together. Reflect on and summarize in 1-2 sentences the different steps taken through this assignment. Compare your conclusions to the findings discussed by the researchers in the article (cite their findings).

**Our analysis aimed to identify the closet species to the fossil specimen GKZ-N 00001 by using clustering and visualization and the article focused on the discovery of the oldest record of the Steller sea lion. We also focused on species classification and visualization but this difers from the paleontological findings in the article. We found the fossil to be close to the Z. japonicus species but the article found it to be the E. jubatus species.**

------------------------------------------------------------------------

## Formatting: (2 pts)

Comment your code, write full sentences, and knit your file!

------------------------------------------------------------------------

```{r, echo=F}
## DO NOT DELETE THIS BLOCK!
Sys.info()
```
