---
title: "Lab 6"
output: html_document
editor_options: 
  chunk_output_type: inline
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE, error = TRUE,
                      warning = FALSE, message = FALSE, 
                      fig.align = "center",
                      R.options = list(max.print=100))

# Edit the file starting below
```

### Andrew Manrriquez

**This assignment is due by the end of the lab. Only one student in the group submits a pdf file on Gradescope.**

*For all questions, include the R commands/functions that you used to find your answer (show R chunk). Answers without supporting code will not receive credit. Write full sentences to describe your findings.*

In this lab, you will explore the `who` dataset which comes with `tidyr`. Let's first load the `tidyverse` package that contains `tidyr`, `dplyr` and `ggplot2` which we will need today:

```{r, warning=FALSE}
# If working on your own computer, remember to install new packages with install.packages("name") in the console
library(tidyverse)
```

Then let's take a quick look at the dataset:

```{r}
# Take a quick look
as_tibble(who)
```

The `who` dataset contains information about tuberculosis (TB) cases per country over the years. The TB cases are reported in the columns `new_sp_m014:newrel_f65` following these conventions:

1.  The first three letters of each column denote `new` cases.

2.  The next two/three letters describe the type of TB: `rel` = cases of relapse, `ep` = cases of extrapulmonary TB, `sn` = cases of pulmonary TB that could not be diagnosed by a pulmonary smear, `sp` = cases of pulmonary TB that could be diagnosed be a pulmonary smear.

3.  The next letter indicates the gender category: females `f` or males `m`.

4.  The remaining numbers gives the age group (for example, 014 means 0-14 years old).

The goal of the lab is to compare tuberculosis (TB) cases across countries over time, first tidying the data.

------------------------------------------------------------------------

### Question 1: (3 pts)

Consider the `who` dataset. Using  `tidvyerse` functions, convert the dataset into a long format by collecting all of the `new_diagnosis_genderage`-columns into a single column (call it `new_diagnosis_genderage`) and all of their values into another single column (call it `cases`). Save the resulting dataset as an R object called `long_who`. How many rows does the `long_who` dataset have?

```{r}
# Converting who dataset into long format
long_who <- who %>%
  pivot_longer(cols = starts_with("new"),
               names_to = "new_diagnosis_genderage",
               values_to = "cases")
nrow(long_who)

```

**The resulting `long_who` data set has 405440 rows.**

------------------------------------------------------------------------

### Question 2: (4 pts)

Next, we would like to separate the `new_diagnosis_genderage` variable into `diagnosis`, `gender`, and `age`. Let's use the `separate()` function. However, there is a discrepancy in coding for new cases of relapse: they appear as `newrel` while all other types of cases use an `_` after `new`. The `mutate` function below will take care of that.

Apply `separate()` to split the variable `new_diagnosis_genderage` into 3 variables called `new`, `diagnosis`, and `genderage`. Then apply `separate()` again to split `genderage` into 2 variables called `gender` and `age`. You will need to specify the separator this time as there is no special character! *Hint: use `sep =` a certain number which corresponds to the number of characters after which to make the split.* Finally, drop the variable `new` since we only have new cases. Call the resulting dataset `tidy_who`. How many columns are left in `tidy_who`?

```{r}
tidy_who <- 
    long_who %>%
    mutate(new_diagnosis_genderage = 
               str_replace(new_diagnosis_genderage, "newrel", "new_rel")) %>%
  separate(col = new_diagnosis_genderage, into = c("new", "diagnosis", "genderage")) %>%
  separate(col = genderage, into = c("gender", "age"), sep = 1) %>% select(-new)

ncol(tidy_who)
    
```

**The `tidy_who` data set has eight columns.**

------------------------------------------------------------------------

### Question 3: (2 pts)

Let's take a look at missing values. The main concern is whether a missing value means that there were no cases of TB or whether it means that the WHO does not have data on the number of TB cases. Are there any zeros in our `tidy_who` dataset? What does it mean about `NA` values?

```{r}
# Check for zeros in the dataset
zeros <- tidy_who %>%
  filter(cases == "0")
# Displays all zero cases
zeros
 
```

**There are zeros in the data set, meaning that the `NA` values are due to a lack of data and not a lack of cases. If `NA` meant that there were no TB cases, then there wouldn't be values with a zero. Therefore, the NA values represent a lack of data collected.**

------------------------------------------------------------------------

### Question 4: (3 pts)

What about missing years for some countries? Those would not appear explicitly in the dataset, they just would not be there. Using `dplyr` functions, find the total number of distinct years for each country in `tidy_who`. 

Also report the minimum and maximum year contained in the dataset for each country. How many countries had less than the expected 34 years (1980 to 2013)? Why do you think these years are missing? *Note: To understand why we have missing years, look at Serbia & Montenegro. What happened to this country?*

```{r}
# Group by country and find distinct years for each country
year_summary <- tidy_who %>%
  group_by(country) %>%
  summarise(distinct_years = n_distinct(year),
            min_year = min(year, na.rm = TRUE),
            max_year = max(year, na.rm = TRUE)) %>%
  ungroup()
# Shows distinct years for each country
year_summary

#Find countries less than the expected 34 years
less_than_34 <- year_summary %>% filter(max_year - min_year + 1 < 34)
# Shows every country with less than 34 years
less_than_34
nrow(less_than_34)

```

**The total number of distinct years for each country can be seen in `year_summary`. There are nine countries with less than 34 years of data. These countries might have missing years due to geopolitical events such as the event that happened with Serbia and Montenegro.**

------------------------------------------------------------------------

### Question 5: (2 pts)

Investigate how the total number of TB cases (adding cases across all diagnoses) changes over the years splitting by gender or age categories. Write a research question/hypothesis to explore this change.

**Research Question: How does the total number of tuberculosis cases change over numerous years, is that change impacted by gender or age categories?**
**Hypothesis: We hypothesize that the total number of TB cases varies over the years potentially being influenced by both gender and age categories. However, we expect to observe differences in TB cases between genders and age categories with certain years displaying higher or lower tuberculosis cases. The fluctuation in TB cases over certain years may be attributed to changing demographics or improvements in reporting.**

------------------------------------------------------------------------

### Question 6: (4 pts)

To answer your research question, first find the annual total number of TB cases in `tidy_who` per age or gender category. Then use a `ggplot` visualization. Include a headline to your visualization and interpret what you see! *Note: remember to handle missing values!*

```{r}
# Calculate annual total TB Cases per age or gender
tb_cases_per_year <- tidy_who %>%
  filter(!is.na(age) & !is.na(gender) & !is.na(year) & !is.na(cases)) %>%
  group_by(year, age, gender) %>%
  summarise(total_cases = sum(cases, na.rm = TRUE))

# Creating ggplot
tb_cases_per_year %>% ggplot(aes(x = year, y = total_cases, color = age)) + 
  geom_line() + 
  facet_wrap(vars(gender)) + 
  labs(title = "Annual Total Tuberculosis Cases by Age & Gender", x = "Year", y = "Total TB Cases", color = "Age Cateogry") + 
  theme_minimal()
```

**Based on the plot we can see that over the years the total TB cases increased for both genders and for all age categories. We see a spike TB cases in both females and males in the 35-44 age category around 2008 - 2009. Around 2015 we see that the total TB cases decreased for both genders and all age categories. We can also see that males tend to contribute more to the total number of TB cases.**

------------------------------------------------------------------------

### Formatting: (2 pts)

Comment your code, write full sentences, and knit your file!
