---
title: "Lab 9"
output: html_document
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE,  
                      warning = FALSE, message = FALSE, 
                      fig.align = "center",
                      R.options = list(max.print=100))

# Edit the file starting below
```

### Andrew Manrriquez

**This assignment is due by the end of the day. Only one student in the group submits a pdf file on Gradescope.**

*For all questions, include the R commands/functions that you used to find your answer (show R chunk). Answers without supporting code will not receive credit. Write full sentences to describe your findings.*

Let's load the appropriate packages for today:

```{r, warning=FALSE}
library(tidyverse)
```

The dataset for this lab contains information about the chances of admissions into graduate school:

```{r}
# Upload data from GitHub
admissions <- read_csv("https://raw.githubusercontent.com/laylaguyot/datasets/main//Admission_Predict.csv") %>%
  # Rename variables for easier manipulation 
  rename(GRE_Score = `GRE Score`, TOEFL_Score = `TOEFL Score`,
         University_Rating = `University Rating`, Admission_Chance = `Chance of Admit`)

# Look at a sample of this data
admissions %>%
  sample_n(size = 10)
```

This dataset contains the following variables: GRE Scores (out of 340), TOEFL Scores (out of 120), 
University Rating (out of 5), Statement of Purpose (`SOP`) and Letter of Recommendation (`LOR`) Strength (out of 5), Undergraduate GPA (out of 10), Research Experience (either 0 or 1), Chance of Admission (ranging from 0 to 1).

The goal of the lab is to make predictions for graduate admission based on other features of a student's application.

------------------------------------------------------------------------

### Question 1: (2 pts)

Pick one variable to predict a student's `Admission_Chance`. What potential relationship between this variable and a student's `Admission_Chance` do you anticipate? Answer that question before exploring the data!

**We predict that there might be a positive relationship between a student's GRE score and that student's admission chances.**

------------------------------------------------------------------------

### Question 2: (4 pts)

Visualize the relationship between the predictor you chose in the previous question and the response variable. Does your visualization match the relationship that you had anticipated?

```{r}
# Creating a scatter plot between GRE scores and admission chance
admissions %>%
  ggplot(aes(x = GRE_Score, y = Admission_Chance)) +
  geom_point() +
  labs(x = "GRE Scores", y = "Admission Chance") +
  ggtitle("Relationship between GRE Scores and Admission Chance")

```

**The visualization indicates that there is a positive relationship between GRE scores and a student's admission chances, which matches our anticipated relationship.**

------------------------------------------------------------------------

### Question 3: (4 pts)

Fit a linear regression model to predict a student's `Admission_Chance` based on the predictor you chose in the first question. Write the expression of the linear model and report the performance of this model.

```{r}
# Fitting a linear regression model to GRE scores and Admission chance
model1 <- lm(Admission_Chance ~ GRE_Score, data = admissions)
summary(model1)
```

**The expression would be E[Admission_Chance] = -2.4360842 + 0.0099759(GRE_Score) and the performance of this model can be determined by the multiple R-squared value. The R-squared value is around .6442 which essentially means that the model explains 64.42% of the variability in Admission_Chance.**

------------------------------------------------------------------------

### Question 4: (4 pts)

Now choose two other predictors to estimate a student's `Admission_Chance`. Visualize the relationship between each predictor you just chose and the response variable. Does there appear to be a relationship between each predictor and the response variable?

```{r}
# Visualize the relationship between TOEFL_Score and Admission_Chance
admissions %>%
  ggplot(aes(x = TOEFL_Score, y = Admission_Chance)) +
  geom_point() +
  labs(x = "TOEFL Scores", y = "Admission Chance") +
  ggtitle("Relationship between TOEFL Scores and Admission Chance")

# Visualize the relationship between CGPA and Admission_Chance
admissions %>%
  ggplot(aes(x = CGPA, y = Admission_Chance)) +
  geom_point() +
  labs(x = "Undergraduate GPA", y = "Admission Chance") +
  ggtitle("Relationship between Undergraduate GPA and Admission Chance")

```

**For both TOEFL scores and undergraduate GPA there appears to be a positive relationship with a student's chances of admission.**

------------------------------------------------------------------------

### Question 5: (4 pts)

Fit a linear regression model to predict a student's `Admission_Chance` based on the two predictors you chose in the previous question. Is this model better than the one with one predictor? Why or why not?

```{r}
# Fitting a linear regression model based on our two predictors
model2 <- lm(Admission_Chance ~ TOEFL_Score + CGPA, data = admissions)
summary(model2)
```

**This model with two predictors is better than the model with one predictor. The second model has a multiple R-squared value of .7774 which means that the model explains 77.74% of the variability in Admission_Chance. This is larger than the first model which only explains 64.42% of the variability in Admission_Chance. The second model also has an RMSE value that is smaller than the RMSE value of the first model.**

------------------------------------------------------------------------

### Formatting: (2 pts)

Comment your code, write full sentences, and knit your file!
